### 2022-05-24 Manuscript/Manuscript.tex
Manuscript updates, stricter schema

\end{flushleft}
\section*{
Abstract
}
The skill to recognize and classify taxa requires knowledge that is less and less common in the scientific community. At the same time, it is clear that these skills are needed in biodiversity monitoring for management and conservation, especially when carried out by citizen scientists. Formalizing the required knowledge in digital identification keys is a way of making such knowledge available for professional and amateur observers of biodiversity. In this paper we describe Clavis, a modern open framework for capturing knowledge required for taxon identification, allowing for a level of detail beyond that of any current format. We exemplify each concept using Pokémon as a fictional taxonomic group.
\linenumbers
\section*{
Introduction
}
Distinguishing biological taxa from one another is a necessity in biodiversity monitoring. As species are going extinct at an unprecedented rate~\cite{Ceballos2015, Johnson2017}, we need to monitor as much as we can to identify statuses and trends. Meanwhile, research and management is currently facing a ``taxonomic impediment'', where taxonomic knowledge is disappearing from the scientific community~\cite{Engel2021}.

@@

Digital identification keys can address these challenges in a number of ways. The formalized and machine-readable way in which the data are stored allows for easier revision, as well as multiple options for their display to the end user, tailored to their level of experience. Digital keys allow for the representation of more complex relationships between species' characteristics and identifications inherent to biological complexity. These relationships cannot as easily be represented in a linear form as is used in traditional, paper-based keys. With a suitable interface, they open up the cumulative experience from taxonomic experts to a broader public like citizen scientists, while being more reliable and educational than other identification methods such as automated image recognition. In contrast to paper-based keys, digital keys can also be combined with one another, integrating the information from several separate keys into one seamless user experience. And conversely, a digital key can easily be limited to just a subset of the taxa in it, for instance only taxa belonging to a certain habitat or recorded in a certain geographic range. Another benefit is the option of saving the user input and the key together with the observational data, as a means of transparent identification that can be reviewed at any time in the future for quality control or in the case of new taxonomical insights.

@@

An alternative approach is the multiple access key. Such keys are typically stored as matrices where taxa and questions (characters) are stored as rows and columns in a tabular format, with cell values linking the taxa to their characters. Each character may have two or more possible answers (states), or be defined as a numerical value or range of values. This allows for a selection of taxa to be made that one wishes to distinguish between, rather than having to traverse the entire key. In a fully populated matrix key, every character is assigned a value (i.e. is scored) for every taxon. The user then has access to all the questions/characters at once and is not required to answer them in any particular order. Choosing any alternative for any of the questions will exclude from consideration all those taxa that are not compatible with that alternative, bringing the user a step closer to the answer. The upside of this approach is that the user has several paths to the answer, and can choose to avoid questions that are difficult or impossible to answer. The downside is that the number of choices can easily be overwhelming, and it is left to the user to try to choose the best path. Also, in real life situations it is very rare to find a set of characters that are both possible to score in a meaningful way for all the taxa in the key, and sufficient for distinguishing between them. More often, some characters will be possible to score for all taxa, whereas others may be essential for distinguishing certain taxa in the key, yet be totally inapplicable to others.

@@

The sparse matrix approach is more powerful than both the single access and the full matrix approach, in that both of the latter approaches are subsets of what the sparse matrix can represent. Still, there is much information one would ideally like to store about taxa, characters, and the relationships between the two, which cannot be easily represented in a tabular format. For instance, the tabular format is not well suited for dealing with polymorphism, i.e. the situation where a taxon can have more than one possible value for a given character. Also, it may often be desirable to treat taxa hierarchically rather than as a simple list of equally ranked entities, something which is difficult to do in the matrix format. Thus, representing taxonomic knowledge in this tabular format restricts the complexity of the information that can be stored.

@@

Clavis compliant keys are written in JSON (JavaScript Object Notation)~\cite{JSON}, which has become the \textit{de facto} standard for data exchange over the internet, as it is an open, flexible format with widespread use and support within all modern programming languages. Clavis itself is a JSON-schema~\cite{JSON_schema}, a formal definition of the structure and content of a valid Clavis JSON file. JSON-schemas are machine readable and can be used to automatically validate the compliance of a JSON file, highlighting any issues that need to be solved. While possible, it is not practical to manually write the JSON of a Clavis-compliant key. For this, key editors will have to be made that provide the taxonomist with an interface that lets them easily record their knowledge, and let the key editor generate the JSON. The latest version of Clavis can be found at \url{https://github.com/Artsdatabanken/Clavis}.

@@

Our aim is for Clavis to serve as an example of how taxonomic knowledge needed for identification might be stored in a way that allows for the representation of the complexity and nuances inherent to such knowledge. The open exchange of taxonomic knowledge, unambiguously captured with as much of the auxiliary details needed for its application, is essential for the preservation of invaluable, increasingly elusive knowledge.

We believe that the storage of taxonomic knowledge with the level of detail exemplified here, combined with user interfaces making it accessible, is vital in enabling observers to gather the data needed for apt nature management. Particularly within citizen science, the potential for tools building on an identification key format such as Clavis is considerable. The more accessible this expert knowledge is, the more accurate the identifications made by the user, and keys on generally less well-known taxa can aid in closing the taxonomic gaps in the data corpus. When storing the user input together with a reported observation, this provides important metadata on the identification and its quality. The result of these projected advances in data collection quality feed back into the areas where these data are used, from research and spatial distribution models to the decision making processes related to the biodiversity crisis in a changing world.

@@

Pokémon and its characters are the intellectual property of The Pokémon Company International. The depictions of Pokémon characters in this manuscript are hand drawn fan art by the authors. Both fan art and scientific use constitute fair use in this context. While reuse is governed by the license under which this paper is published, use of these images in contexts that do not constitute fair use may infringe on trademark rights of The Pokémon Company International. As stated in the license under which this manuscript is published, the authors bear no responsibility for any use by others, wrongful or not, and recommend consulting with legal experts or The Pokémon Company International prior to reuse to avoid dispute.

>>>>

\end{flushleft}
\section*{
Abstract
}
The skill to recognize and classify taxa requires knowledge that is less and less common in the scientific community. At the same time, it is clear that these skills are needed in biodiversity monitoring for management and conservation, especially when carried out by citizen scientists. Formalizing the required knowledge in digital identification keys is a way of making such knowledge more available for professional and amateur observers of biodiversity. In this paper we describe Clavis, a modern open framework for capturing knowledge required for taxon identification through digital keys, allowing for a level of detail beyond that of any current format. We exemplify each concept using Pokémon as a fictional taxonomic group.
\linenumbers
\section*{
Introduction
}
Distinguishing biological taxa from one another is a necessity in biodiversity monitoring. As species are going extinct at an unprecedented rate~\cite{Ceballos2015, Johnson2017}, we need to monitor as much as we can to identify statuses and trends. Meanwhile, research and management is currently facing a ``taxonomic impediment'', where taxonomic knowledge is disappearing from the scientific community~\cite{Engel2021}.

@@

Digital identification keys can address these challenges in a number of ways. The formalized and machine-readable way in which the data are stored allows for easier revision, as well as multiple options for their display to the end user, tailored to their level of experience. Digital keys allow for the representation of more complex relationships between species' characteristics and identifications inherent to biological complexity. These relationships cannot as easily be represented in a linear form as is used in traditional, paper-based keys. With a suitable interface, they open up the cumulative experience from taxonomic experts to a broader public like citizen scientists, while potentially being more reliable and educational than other identification methods such as automated image recognition. In contrast to paper-based keys, digital keys can also be combined with one another, integrating the information from several separate keys into one seamless user experience. And conversely, a digital key can easily be limited to just a subset of the taxa in it, for instance only taxa belonging to a certain habitat or recorded in a certain geographic range. Another benefit is the option of saving the user input and the key together with the observational data, as a means of transparent identification that can be reviewed at any time in the future for quality control or in the case of new taxonomical insights.

@@

An alternative approach is the multiple access key. Such keys are typically stored as matrices where taxa and questions (characters) are stored as rows and columns in a tabular format, with cell values linking the taxa to their characters. Each character may have two or more possible answers (states), or be defined as a numerical value or range of values. This allows for a selection of taxa to be made that one wishes to distinguish between, rather than having to traverse the entire key. In a fully populated matrix key, every character is assigned a value (i.e. is scored) for every taxon. The user then has access to all the questions/characters at once and is not required to answer them in any particular order. Choosing any alternative for any of the questions will exclude from consideration all those taxa that are not compatible with that alternative, bringing the user a step closer to the answer. The upside of this approach is that the user has several paths to the answer, and can choose to avoid questions that are difficult or impossible to answer. The downside is that the number of choices can easily be overwhelming, and it is generally left to the user to try to choose the best path. Also, in real life situations it is very rare to find a set of characters that are both possible to score in a meaningful way for all the taxa in the key, and sufficient for distinguishing between them. More often, some characters will be possible to score for all taxa, whereas others may be essential for distinguishing certain taxa in the key, yet be totally inapplicable to others.

@@

The sparse matrix approach is more powerful than both the single access and the full matrix approach, in that both of the latter approaches are subsets of what the sparse matrix can represent. Still, there is much information one would ideally like to store about taxa, characters, and the relationships between the two, which cannot be easily represented in a tabular format. For instance, the tabular format is not well suited for dealing with polymorphism in a straightforward way, i.e. the situation where a taxon can have more than one possible value for a given character. Also, it may often be desirable to treat taxa hierarchically rather than as a simple list of equally ranked entities, something which is difficult to do in the matrix format. Thus, representing taxonomic knowledge in this tabular format restricts the complexity of the information that can be stored.

@@

Clavis compliant keys are written in JSON (JavaScript Object Notation)~\cite{JSON}, which has become the \textit{de facto} standard for data exchange over the internet, as it is an open, flexible format with widespread use and support within all modern programming languages. Clavis itself is a JSON-schema~\cite{JSON_schema}, a formal definition of the structure and content of a valid Clavis JSON file. JSON-schemas are machine readable and can be used to automatically validate the compliance of a JSON file, highlighting any issues that need to be solved. While possible, it is generally not practical to manually write the JSON of a Clavis-compliant key. In order to facilitate key creation and editing, one would generally provide taxonomists with a key editing interface that lets them easily record their knowledge, and let the key editor generate the JSON. The latest version of Clavis can be found at \url{https://github.com/Artsdatabanken/Clavis}.

@@

Our aim is for Clavis to be a relevant tool for storing taxonomic knowledge needed for identification in a way that allows for the representation of the complexity and nuances inherent to such knowledge. The open exchange of taxonomic knowledge, unambiguously captured with as much of the auxiliary details needed for its application, is essential for the preservation of invaluable, increasingly elusive knowledge.

We believe that the storage of taxonomic knowledge with the level of detail exemplified here, combined with user interfaces making it accessible, is vital in enabling observers to gather the data needed for apt nature management. Particularly within citizen science, the potential of tools built on an identification key format such as Clavis is considerable. The more accessible this expert knowledge is, the more accurate the identifications made by the user, and keys on generally less well-known taxa can aid in closing the taxonomic gaps in the data corpus. When storing the user input together with a reported observation, this provides important metadata on the identification and its quality. The result of these projected advances in data collection quality feed back into the areas where these data are used, from research and spatial distribution models to the decision making processes related to the biodiversity crisis in a changing world.

@@

Fig. \ref{fig5} is based on an Odonata key previously published by the Norwegian Biodiversity Information Centre. In these screenshots, character illustrations are made by Hallvard Elven and licensed under a CC BY 4.0 license, while taxon images are made by Göran Liljeberg and licensed under a CC BY-SA 4.0 license.

Pokémon and its characters are the intellectual property of The Pokémon Company International. The depictions of Pokémon characters in this manuscript are hand drawn fan art by the authors. Both fan art and scientific use constitute fair use in this context. While reuse is governed by the license under which this paper is published, use of these images in contexts that do not constitute fair use may infringe on trademark rights of The Pokémon Company International. As stated in the license under which this manuscript is published, the authors bear no responsibility for any use by others, and recommend consulting with legal experts or The Pokémon Company International prior to reuse to avoid dispute.
